---
title: "Therapy goals and evaluation"
output:
  pdf_document:

    fig_caption: true
    df_print: kable
  word_document: 
    fig_caption: true
    df_print: kable
  number_sections: true
url_colour: blue
---


<!-- Global settings and Packages-->
```{r, results='hide', echo = FALSE, message=FALSE}
# settings to hide code, messages and warnings in PDF output
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
knitr::opts_knit$set(root.dir = '/Users/Hannah/Documents/Analyse_Julia_Sonnleitner/pain/' ) #setting all root directories relative to this one

#Packages
library(RColorBrewer)
library(dplyr) #for basic data handling
library(ggplot2) #for figures
library(scales)
library(ggpubr)
library(gridExtra) #for multiple figures
Sys.setlocale("LC_ALL", 'en_US.UTF-8')
options(scipen=10)
```

```{r, prepare data, results=FALSE}
dge <- read.csv2('../data/therapy_goals_and_evaluation.csv') %>% 
  mutate(group = factor(group, labels = c("Treatment"))) 

#missing values per question
apply(apply(dge[,c(3:16, 19:32, 35)], 2, is.na), 2, sum)
#which subjects
filter(dge, is.na(knowledge_pain))

d <- dge %>% 
  filter(!is.na(knowledge_pain))

summary(d)
```

# Missing values

2 patients have missing values for all questions, these are omitted from all plots. One patient has a missing value for the question on confidence about future progress, this missing value is explicitly shown in the plot for this question. 

```{r, figure settings, results=FALSE}
# figure settings: ####
increasing <- brewer.pal(8,"YlGnBu")[c(4:8)]#colours from set YlGnbu = increasing blue tones
st = 15 # font size title
sl = 12 #label font size

```


# Therapeutic goals: In how far did therapy help to achieve positive changes? 

Patients rated the extent of achieved change as: 

* 5 = strongly improved
* 4 = clearly improved
* 3 = partially improved 
* 2 = improved a bit
* 1 = did not improve/unchanged

```{r, single goal plots, results=FALSE, warnings=FALSE}
#function to add all settings to the plot for goal questions
plotgoal <- function(plot){
  plot +
    geom_histogram(stat="count")+
    scale_y_continuous(breaks=c(0, 2,4,6, 8), name="Count", limits=c(0,9))+
    scale_x_discrete(breaks = c(1:5, NA), labels = c("no change", "a bit", "partially", "clearly", "strongly", "missing"))+
    scale_fill_manual(values=increasing, na.value="grey")+
    theme_bw() +
    theme(axis.title.x= element_blank(), text=element_text(size=sl),plot.title=element_text(size=sl), 
          panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          legend.position="none", plot.margin = unit(rep(0.5, 4), "cm"))
}

plot <- ggplot(d, aes(x=factor(knowledge_pain),fill=factor(knowledge_pain)))+
  ggtitle("Knowledge about chronic pain")
p1 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(knowledge_med),fill=factor(knowledge_med)))+
  ggtitle("Knowledge about medicine/therapies")
p2 <- plotgoal(plot)


plot <- ggplot(d, aes(x=factor(muscle_relax),fill=factor(muscle_relax)))+
  ggtitle("Muscle relaxation")
p3 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(stressors),fill=factor(stressors)))+
  ggtitle("Recognizing pain triggers/stressors")
p4 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(guide_attention),fill=factor(guide_attention)))+
  ggtitle("Diverting attention")
p5 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(focus_pleasure),fill=factor(focus_pleasure)))+
  ggtitle("Focusing on something pleasant")
p6 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(treat_self),fill=factor(treat_self)))+
  ggtitle("Treat myself with beautiful/new things")
p7 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(stop_negthoughts),fill=factor(stop_negthoughts)))+
  ggtitle("Stop negative/promote positive thoughts")
p8 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(wellbeing),fill=factor(wellbeing)))+
  ggtitle("Promoting my subjective wellbeing")
p9 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(conflicts),fill=factor(conflicts)))+
  ggtitle("Coping with problems or conflicts")
p10 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(reduce_pain_intensity),fill=factor(reduce_pain_intensity)))+
  ggtitle("Reducing pain intensity")
p11 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(reduce_pain_duration),fill=factor(reduce_pain_duration)))+
  ggtitle("Reducing pain duration")
p12 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(pain_control),fill=factor(pain_control)))+
  ggtitle("Controlling the pain/other complaints")
p13 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(lq),fill=factor(lq)))+
  ggtitle("Improving life quality")
p14 <- plotgoal(plot)

plot <- ggplot(d, aes(x=factor(other),fill=factor(other)))+
  ggtitle("Other areas of change")
p15 <- plotgoal(plot)
```

\newpage

```{r,  goal figures 1, fig.width=7, fig.height = 9}
grid.arrange(p1, p2, p3, p4, p5, p6, ncol=2)
```

\newpage

```{r,  goal figures 2, fig.width=7, fig.height = 9, cache = FALSE}
grid.arrange(p7, p8, p9, p10, p11, p12, ncol=2)

```

\newpage

```{r, goal figures 3,  fig.width=7, fig.height = 6}

grid.arrange(p13, p14, p15, ncol=2)

```

3 patients described other areas of change in words: 

* Wissen dass man nicht alleine ist
* Dass Andere noch groessere Probleme haben  
* Viel aktiver   

\newpage 

## Therapy goals: Medians for each item

```{r, goals median figures}
d_median = d %>% 
  select(-c(other, subid)) %>% #exclude text entry variables, and subids
  select_if(is.integer) %>% # exclude factors
  summarize_all(median, na.rm = TRUE) %>% 
  gather(key='item', value='median') %>% 
  mutate(variable=c(rep('goals', 14), rep('evaluation', 15)),
         item_text=c("Knowledge about chronic pain", "Knowledge about medicine/therapies",
                     "Muscle relaxation", "Recognizing pain triggers/stressors", "Diverting attention",
                     "Focusing on something pleasant","Treat myself with beautiful/new things",
                     "Stop negative/promote positive thoughts", "Promoting my subjective wellbeing", 
                     "Coping with problems or conflicts", "Reducing pain intensity", "Reducing pain duration", 
                     "Controlling the pain/other complaints", "Improving life quality",
                     "I had positive expectations about GT.", "I could talk openly in the group.",
                     "I participated actively.", "The behavior of group leader was\nhelpful.",
                     "Sessions were well planned and\nstructured.", "The group leader was understanding.",
                     "After this GT, I am confident that\nI will make further progress.",
                     "The GT increased trust in myself.", "The GT was too much for me.",
                     "GT did not help for my problems.","I found the GT helpful.",
                     "I felt well in the group.",  "Other group members disturbed me.",
                     "Single therapy sessions would\nhave helped more.", "In total, I am satisfied with the GT."),
         median = factor(median, ordered=TRUE))

#plot
pgoalsmedian =ggplot(filter(d_median, variable=='goals'), aes(x=item_text, y=median, width=0.6, fill=variable))+
  geom_bar(stat="identity")+coord_flip()+
  scale_y_discrete(breaks = c(1:5), labels = c("no change", "a bit", "partially", "clearly", "strongly"), drop=FALSE)+
  scale_fill_manual(values=c(increasing[3]))+
  theme_bw() +
  theme(axis.title= element_blank(), text=element_text(size=sl),plot.title=element_text(size=sl), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position="none", plot.margin = unit(rep(0.5, 4), "cm"))+
  ggtitle('Therapy goals (Median)')
pdf('../figures/therapy_goals_median.pdf'); pgoalsmedian; dev.off()
```


# Evaluation of group therapy

Patients rated in how far they agreed to each statement. GT is the group therapy. 

```{r, single evaluation plots, results=FALSE, warnings=FALSE}
#function to add all settings to the plot for goal questions
ploteval <- function(plot){
  plot +
    geom_histogram(stat="count")+
    scale_y_continuous(breaks=c(2, 4,6,8,10, 12), limits=c(0,12), name="Count")+
    scale_x_discrete(breaks = c(1:5, NA), labels = c("not\naccurate", "not very\naccurate", "partially\naccurate", "pretty\naccurate", "totally\naccurate", "missing"))+
    scale_fill_manual(values=increasing, na.value="grey")+
    theme_bw() +
    theme(axis.title.x= element_blank(), text=element_text(size=sl),plot.title=element_text(size=sl), 
          panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          legend.position="none", plot.margin = unit(rep(0.5, 4), "cm"))
}


plot <- ggplot(d, aes(x=factor(had_positive_expectations),fill=factor(had_positive_expectations)))+
  ggtitle("I had positive expectations about GT.")
p1 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(could_talk_openly),fill=factor(could_talk_openly)))+
  ggtitle("I could talk openly in the group.")
p2 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(I_participated_actively),fill=factor(I_participated_actively)))+
  ggtitle("I participated actively.")
p3 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(group_leader_helpful),fill=factor(group_leader_helpful)))+
  ggtitle("The behavior of group leader was\nhelpful.")
p4 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(session_structure_planning_good),fill=factor(session_structure_planning_good)))+
  ggtitle("Sessions were well planned and\nstructured.")
p5 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(leader_was_understanding),fill=factor(leader_was_understanding)))+
  ggtitle("The group leader was understanding.")
p6 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(confident_future_progress),fill=factor(confident_future_progress)))+
  ggtitle("After this GT, I am confident that\nI will make further progress.")
p7 <- ploteval(plot)+
  theme(axis.title.x= element_blank(), text=element_text(size=10),plot.title=element_text(size=sl), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position="none", plot.margin = unit(rep(0.5, 4), "cm"))

plot <- ggplot(d, aes(x=factor(increase_trust),fill=factor(increase_trust)))+
  ggtitle("The GT increased trust in myself.")
p8 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(therapy_toomuch),fill=factor(therapy_toomuch)))+
  ggtitle("The GT was too much for me.")
p9 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(therapy_nohelp),fill=factor(therapy_nohelp)))+
  ggtitle("GT did not help for my problems.")
p10 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(therapy_helpful),fill=factor(therapy_helpful)))+
  ggtitle("I found the GT helpful.")
p11 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(felt_well_in_group),fill=factor(felt_well_in_group)))+
  ggtitle("I felt well in the group.")
p12 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(others_disturbed_me),fill=factor(others_disturbed_me)))+
  ggtitle("Other group members disturbed me.")
p13 <- ploteval(plot)

plot <- ggplot(d, aes(x=factor(prefer_therapy_alone),fill=factor(prefer_therapy_alone)))+
  ggtitle("Single therapy sessions would\nhave helped more.")
p14 <- ploteval(plot)

p15 <- ggplot(d, aes(x=(total),fill=factor(total)))+
  ggtitle("In total, I am satisfied with the GT.")+
  geom_histogram(stat="count")+
  scale_y_continuous(breaks=c(2,4,6,8,10,12), limits=c(0,12), name="Count")+
  scale_x_continuous(breaks = c(0:10), limits=c(0,10.5), labels = c("not at\nall", "1", "2", "3", "4", "5", "6", "7", "8", "9", "totally"))+
  scale_fill_manual(values=increasing, na.value="grey")+
  theme_bw() +
  theme(axis.title.x= element_blank(), text=element_text(size=sl),plot.title=element_text(size=sl), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position="none", plot.margin = unit(rep(0.5, 4), "cm"))
```


Patients evaluated the group therapy (GT) on the following scale: 

* 5 = totally accurate
* 4 = pretty accurate
* 3 = partially accurate
* 2 = not very accurate
* 1 = not accurate

\newpage

```{r,  evaluation figures 1, fig.width=7, fig.height = 9}
grid.arrange(p1, p2, p3, p6, p4, p5, ncol=2)
```

\newpage

```{r,  evaluation figures 2, fig.width=7, fig.height = 9, cache = FALSE}
grid.arrange(p8, p9, p10, p11, p12, p13, ncol=2)
```

\newpage

```{r, evaluation figures 3,  fig.width=7, fig.height = 3}
grid.arrange(p7, p14,  ncol=2)
```

--

```{r, total evaluation score, fig.width=4.5, fig.height = 3}
p15
```

## Therapy evaluation: Median for each item

```{r}
#plot
pevalmedian =ggplot(filter(d_median, variable=='evaluation' & item!="total"), aes(x=item_text, y=median, width=0.6, fill=variable))+
  geom_bar(stat="identity")+coord_flip()+
    scale_y_discrete(breaks=c(1:5), drop=FALSE, 
                      labels = c("not\naccurate", "not very\naccurate", "partially\naccurate", "pretty\naccurate", "totally\naccurate"), name="Count")+
    # scale_x_discrete(breaks = c(1:5, NA), labels = c("not\naccurate", "not very\naccurate", "partially\naccurate", "pretty\naccurate", "totally\naccurate", "missing"))+
  #scale_y_discrete(breaks = c(1:5), labels = c("no change", "a bit", "partially", "clearly", "strongly"), drop=FALSE)+
  scale_fill_manual(values=c(increasing[3]))+
  theme_bw() +
  theme(axis.title= element_blank(), text=element_text(size=sl),plot.title=element_text(size=sl), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position="none", plot.margin = unit(rep(0.5, 4), "cm"))+
  ggtitle('Therapy evaluation (Median)')
pdf('../figures/therapy_evaluation_median.pdf'); pevalmedian; dev.off()
```

